\textbf{A framework for effective application of machine learning to microbiome-based classification problems}

### 2020-02-21 submission/MicrobeConference/Microbe_Abstract2019.Rmd
final edits before resubmission

As gut microbiome field continues to grow, there is an ever-increasing demand for reproducible machine learning methods to determine associations between the microbiome and a phenotype of interest. Currently, the use of machine learning in microbiome research lacks clarity and consistency over the modeling pipeline (training, validation and testing steps). There is a need for guidance on how to implement good machine learning practices to generate reproducible and robust models.

Recently, there has been growing interest in using machine learning to identify colorectal tumors with microbiota-associated biomarkers. Colorectal cancer is one of the leading cause of death among cancers in the United States. Colonoscopy as a screening tool is effective, however it is invasive and have a low rate of patient adherence. Previous studies have shown that bacterial abundances in the stool can predict colorectal tumors in the colon and can be used as a non-invasive screening tool. However, the prediction performance of these models vary greatly, with areas under the receiver operating characteristic curve (AUC) of 0.7-0.9 [@sze_leveraging_2018; @baxter_microbiota-based_2016; @baxter_dna_2016; @zackular_human_2014]. The variation in prediction performance is based in part on differences in the study populations, and in part on the differences in modeling pipelines.

In this study, hemoglobin counts and 16S gene sequences in the stool were used to identify colorectal tumors of `r paste(nrow(meta))` patients as advanced tumors or not. Modeling pipelines were established for L2-regularized Logistic Regression, L1 and L2 Linear Support Vector Machines (SVM), Radial Basis Function SVM, Decision Tree, Random Forest and XGBoost binary classification models. The mean AUCs of these models were `r paste(round(test_meanAUC,2))` ± `r paste(round(test_sdAUC,2))`,  `r paste(round(l1svm_summary[2,2],2))` ± `r paste(round(l1svm_summary[2,3],2))`,  `r paste(round(l2svm_summary[2,2],2))` ± `r paste(round(l2svm_summary[2,3],2))`, `r paste(round(svmRBF_summary[2,2],2))` ± `r paste(round(svmRBF_summary[2,3],2))`, `r paste(round(dt_summary[2,2],2))` ± `r paste(round(dt_summary[2,3],2))`, `r paste(round(rf_summary[2,2],2))` ± `r paste(round(rf_summary[2,3],2))`, and `r paste(round(xgboost_summary[2,2],2))` ± `r paste(round(xgboost_summary[2,3],2))`, respectively. Tree-based methods, namely Decision Tree, Random Forest and XGBoost were less susceptible to overfitting and in general had higher sensitivity and specificity for advanced tumors. Aside from evaluating generalization and classification performance of each model, this study established standards for modeling pipelines of microbiome-associated machine learning models.

>>>>

As the microbiome field continues to grow, there is an ever-increasing demand for reproducible methods for identifying associations between members of the microbiome and phenotypes. Currently, the field’s use of machine learning lacks clarity and consistency. There is a need for guidance on how to implement good machine learning practices to generate reproducible and robust models.

One application of machine learning to microbiome data has been to classify patients as having colorectal tumors based on microbiota-associated biomarkers. Colorectal cancer is one of the leading cause of death among cancers in the United States. Colonoscopy as a screening tool is effective, however it is invasive and have a low rate of patient adherence. Previous studies have shown that bacterial abundances in the stool can predict colorectal tumors in the colon and can be used as a non-invasive screening tool.  However, the prediction performance of these models vary greatly, with areas under the receiver operating characteristic curve (AUC) of 0.7-0.9 [@sze_leveraging_2018; @baxter_microbiota-based_2016; @baxter_dna_2016; @zackular_human_2014]. The variation in prediction performance is based in part on differences in the study populations, and in part on the differences in modeling pipelines.

We used the fecal hemoglobin concentration and 16S rRNA gene sequences from stool samples to classify `r paste(nrow(meta))` patients as having advanced tumors or not. Modeling pipelines were established for L2-regularized Logistic Regression, L1 and L2 Linear Support Vector Machines (SVM), Radial Basis Function SVM, Decision Tree, Random Forest and XGBoost binary classification models. The mean AUCs of these models were `r paste(round(test_meanAUC,2))` ± `r paste(round(test_sdAUC,2))`,  `r paste(round(l1svm_summary[2,2],2))` ± `r paste(round(l1svm_summary[2,3],2))`,  `r paste(round(l2svm_summary[2,2],2))` ± `r paste(round(l2svm_summary[2,3],2))`, `r paste(round(svmRBF_summary[2,2],2))` ± `r paste(round(svmRBF_summary[2,3],2))`, `r paste(round(dt_summary[2,2],2))` ± `r paste(round(dt_summary[2,3],2))`, `r paste(round(rf_summary[2,2],2))` ± `r paste(round(rf_summary[2,3],2))`, and `r paste(round(xgboost_summary[2,2],2))` ± `r paste(round(xgboost_summary[2,3],2))`, respectively. Tree-based methods, namely Decision Tree, Random Forest and XGBoost were less susceptible to overfitting and in general had higher sensitivity and specificity for advanced tumors. Aside from evaluating generalization and classification performance of each model, this study established standards for modeling pipelines of microbiome-associated machine learning models.

### 2020-02-21 submission/MicrobeConference/Microbe_Abstract2019.tex
final edits before resubmission

\linenumbers
\textbf{Evaluation of classification pipelines that predict colorectal
cancer progression with microbiota-associated biomarkers}

Begüm D. Topçuoğlu\({^1}\), Jenna Wiens\({^2}\), Patrick D.
Schloss\textsuperscript{1\(\dagger\)}

As gut microbiome field continues to grow, there will be an
ever-increasing demand for reproducible machine learning methods to
determine association of the microbiome with a continuous or categorical
phenotype of interest. Currently, the use of machine learning in
microbiome literature lack clarity over the training, validation and
testing of the models used. There is a need to properly implement good
machine learning practices to generate reproducible and robust models.

Recently, there is an interest in using machine learning to predict
colorectal cancer progression with microbiota-associated biomarkers.
Colorectal cancer is one of the leading cause of death among cancers in
the United States. Each person in the industrialized world has on
average a one-in-twenty chance of developing colorectal cancer (CRC)
(1--3). Colonoscopy as a screening tool is effective, however it is
invasive, expensive and have a low rate of patient adherence. Therefore,
gut microbiome-based biomarkers emerged as a non-invasive screening
method.

In this study, classification models that use human hemoglobin levels
and bacterial population abundances in the stool were used to predict
colorectal disease status as screen-relevant colonic growth or not.
Training, validation and testing pipelines were established for
L2-regularized Logistic Regression, L1 and L2 Linear Suppor Vector
Machines (SVM), Radial Basis Function SVM, Decision Tree, Random Forest
and XGBoost classifiers. The generalization and prediction performance
of these classifiers were evaluated and each classifier was examined for
its reproducibility, robustness and susceptibility to overfitting.
L2-regularized Logistic Regression had a mean AUC of 0.68 +/- 0.04, L1
Linear SVM had a mean AUC of 0.76 +/- 0.05, L2 Linear SVM had a mean AUC
of 0.68 +/- 0.05 and Radial Basis Function SVM had a mean AUC of 0.69
+/- 0.05. Decision Tree had a mean AUC of 0.7 +/- 0.05, Random Forest
had a mean AUC of 0.76 +/- 0.06 and XGBoost had a mean AUC of 0.76 +/-
0.04. Tree-based models were less susceptible to overfiting and in
general had higher sensitivity and specificity for colonic
screen-relevant growth.

@@

\subsection{References}\label{references}

\hypertarget{refs}{}
\hypertarget{ref-seer_2016}{}
1. \textbf{Howlader N KM \textnormal{Noone AM}}. SEER cancer statistics
review, 1975-2013, (national cancer institute. bethesda, md).

\hypertarget{ref-street_colorectal_nodate}{}
2. \textbf{Street W}. Colorectal cancer facts \& figures 2017-2019 40.

\hypertarget{ref-weir_past_2015}{}
3. \textbf{Weir HK}, \textbf{Thompson TD}, \textbf{Soman A},
\textbf{MÃžller B}, \textbf{Leadbetter S}. 2015. The past, present, and
future of cancer incidence in the united states: 1975 through 2020.
Cancer \textbf{121}:1827--1837.
doi:\href{https://doi.org/10.1002/cncr.29258}{10.1002/cncr.29258}.

>>>>

\linenumbers
\textbf{Evaluation of machine learning methods that identify colorectal
lesions with microbiota-associated biomarkers}

Begüm D. Topçuoğlu, Jenna Wiens, Mack Ruffin, Patrick D. Schloss

As gut microbiome field continues to grow, there is an ever-increasing
demand for reproducible machine learning methods to determine
associations between the microbiome and a phenotype of interest.
Currently, the use of machine learning in microbiome research lacks
clarity and consistency over the modeling pipeline (training, validation
and testing steps). There is a need for guidance on how to implement
good machine learning practices to generate reproducible and robust
models.

Recently, there has been growing interest in using machine learning to
identify colorectal lesions that are precursors of coloractal cancer,
with microbiota-associated biomarkers. Colorectal cancer is one of the
leading cause of death among cancers in the United States. Colonoscopy
as a screening tool is effective, however it is invasive, expensive and
have a low rate of patient adherence. Previous studies have shown that
bacterial population abundances in the stool can predict screen relevant
lesions in the colon and can be used as a non-invasive screening tool.
However, the prediction performance of these models vary greatly, with
areas under the receiver operating characteristic curve (AUC) of 0.7-0.9
(1--4). The variation in prediction performance is based in part on
differences in the study populations, and in part on the differences in
modeling pipelines.

In this study, hemoglobin levels and 16S rRNA gene sequences in the
stool were used to identify colorectal lesions of 490 patients. The
colorectal disease stage was defined as showing screen-relevant lesions
or not. Modeling pipelines were established for L2-regularized Logistic
Regression, L1 and L2 Linear Support Vector Machines (SVM), Radial Basis
Function SVM, Decision Tree, Random Forest and XGBoost binary
classification models. The mean AUCs of these models were 0.68 ± 0.04,
0.76 ± 0.05, 0.68 ± 0.05, 0.69 ± 0.05, 0.71 ± 0.04, 0.82 ± 0.04, and
0.76 ± 0.04, respectively. Tree-based methods, namely Decision Tree,
Random Forest and XGBoost were less susceptible to overfitting and in
general had higher sensitivity and specificity for colonic
screen-relevant lesions. Aside from evaluating generalization and
classification performance of each model, this study established
standards for modeling pipeline of the microbiome-associated machine
learning models.

@@

\textbf{References}

\hypertarget{refs}{}
\hypertarget{ref-sze_leveraging_2018}{}
1. \textbf{Sze MA}, \textbf{Schloss PD}. 2018. Leveraging existing 16S
rRNA gene surveys to identify reproducible biomarkers in individuals
with colorectal tumors. mBio \textbf{9}:e00630--18.
doi:\href{https://doi.org/10.1128/mBio.00630-18}{10.1128/mBio.00630-18}.

\hypertarget{ref-baxter_microbiota-based_2016}{}
2. \textbf{Baxter NT}, \textbf{Ruffin MT}, \textbf{Rogers MAM},
\textbf{Schloss PD}. 2016. Microbiota-based model improves the
sensitivity of fecal immunochemical test for detecting colonic lesions.
Genome Medicine \textbf{8}:37.
doi:\href{https://doi.org/10.1186/s13073-016-0290-3}{10.1186/s13073-016-0290-3}.

\hypertarget{ref-baxter_dna_2016}{}
3. \textbf{Baxter NT}, \textbf{Koumpouras CC}, \textbf{Rogers MAM},
\textbf{Ruffin MT}, \textbf{Schloss PD}. 2016. DNA from fecal
immunochemical test can replace stool for detection of colonic lesions
using a microbiota-based model. Microbiome \textbf{4}.
doi:\href{https://doi.org/10.1186/s40168-016-0205-y}{10.1186/s40168-016-0205-y}.

\hypertarget{ref-zackular_human_2014}{}
4. \textbf{Zackular JP}, \textbf{Rogers MAM}, \textbf{Ruffin MT},
\textbf{Schloss PD}. 2014. The human gut microbiome as a screening tool
for colorectal cancer. Cancer Prev Res \textbf{7}:1112--1121.
doi:\href{https://doi.org/10.1158/1940-6207.CAPR-14-0129}{10.1158/1940-6207.CAPR-14-0129}.
